%! BIB program = biber
\documentclass[12pt,a4paper]{article}

% Pacchetti utili
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{setspace}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[backend=biber,style=numeric]{biblatex}
\usepackage{csquotes}

\addbibresource{references.bib}

\hypersetup{
    colorlinks=true,
    linkcolor=black,
    urlcolor=blue,
    citecolor=black
}

% Margini
\geometry{margin=1in}

% Interlinea
\onehalfspacing

\begin{document}

% Title
\begin{titlepage}
    \centering
    \vspace*{3cm}
    
    {\LARGE \textbf{JOMAL PP}}\\[0.5cm]
    {\large Deep Learning for Art Classification}\\[10cm]
    
    \textbf{Authors:} Andreea Roica, Jenny Cubelo, Libero Biagi, Marisa Esteves, Oliver Kain \\
    \textbf{Student IDs:} 20250361, 20250431, 20250349, 20250348, 20250401 \\
    \textbf{Course:} Deep Learning \\
    \textbf{University:} NOVA Information Management School \\
    \textbf{Date:} \today
    
    \vfill
\end{titlepage}

% ===== INDICE =====
\tableofcontents
\newpage

% ===== SEZIONI =====
\section{Introduction}

The classification of artistic images poses unique challenges compared to conventional image recognition tasks. Paintings often exhibit significant variability in style, composition, and abstraction, even within the same artist class, while distinct classes may share similar visual characteristics. As a result, effective models must capture subtle patterns such as texture, color distribution, and stylistic features that are commonly learned through deep convolutional representations \cite{krizhevsky2012,gatys2016}.

This project addresses the problem of classifying artworks from a subset of the WikiArt dataset using deep learning techniques. The objective is to develop models capable of learning discriminative visual representations that generalize well despite the complexity and variability of the data.

To this end, several deep learning approaches are explored and evaluated, with a focus on understanding how different architectures capture the underlying structure of artistic images. Convolutional neural networks have been shown to be effective at learning hierarchical feature representations from images, while also being applicable to style- and texture-sensitive visual tasks \cite{gatys2016}. The performance of these models is assessed using standard evaluation metrics, providing insight into their effectiveness for this task.


\section{Dataset and Data Preparation}
\subsection{Dataset Description}

The dataset used in this project is a subset of the WikiArt collection, which contains digitized paintings from 23 artists. 
The task considered is artist classification, where each image is associated with the artist who created the work, forming a multi-class classification problem.
The distribution of samples is not perfectly balanced across classes, with the number of images per artist ranging from 336 to 1322, for a total of 13,340 images.
This imbalance introduces additional challenges for model training and evaluation.


\subsection{Data Splitting Strategy}

To ensure that the model generalizes across different visual styles of each artist and to address the class imbalance, we adopted a clustering-based splitting strategy within each artist. 
For each image, we extracted low-level visual features, including a grayscale indicator and statistical descriptors in HSV color space, after resizing images to $64 \times 64$ and normalizing pixel values.
Clustering was performed independently per artist using KMeans, with the number of clusters selected by maximizing the silhouette score. 
This groups artworks with similar visual characteristics (e.g., color palette or style) within each artist.

Initially, we experimented with a purely random split by artist, but observed imbalanced results. 
This led to situations where the model was not exposed during training to certain stylistic variations, negatively affecting performance. 


The final dataset split was obtained by stratifying on the combined artist–cluster label, using an 80/10/10 ratio for training, validation, and testing. 
This ensures that each split contains samples from all clusters of every artist, exposing the model to the full range of styles during training.
The 80/10/10 ratio was chosen to maximize the number of training samples while maintaining sufficiently large and balanced validation and test sets.

Stratification by artist–cluster helps preserve both class distribution and intra-class variability across splits, leading to a more robust and fair evaluation of the classification model.

\subsection{Data Preprocessing}
\subsubsection{Normalization}

For the purpose of normalization, images were decoded into RGB format and pixel values were scaled from the range [0,255] to [0,1], ensuring consistent input representation.
The same preprocessing was applied uniformly to the training, validation, and test sets. only in our net. the other models were pre trained and we used the normalization that they were trained with.

\subsubsection{Resizing}

We analyzed the resolution of all images in the dataset by extracting their dimensions. 
The results showed that all images share a resolution of $512 \times 512$ pixels, which is relatively high for efficient model training. 
Therefore, images were considered for downscaling to a smaller size to reduce computational cost.

\subsection{Data Augmentation}

Techniques used (e.g., rotations, flips, color jitter) and rationale.

\subsection{Batching}

The dataset was processed in batches of 64 samples, as it provides a good balance between memory usage and training performance. 
Each batch consists of tensors of shape (64, 512, 512, 3) for images and (64,) for labels.

\section{Methodology}

\subsection{Custom CNN Architecture}

The proposed CNN follows a design in which the number of filters increases progressively across convolutional blocks (32, 64, 128, 256). 
This allows the network to learn increasingly complex representations: early layers capture low-level features such as edges and colors, while deeper layers encode higher-level patterns such as textures and stylistic characteristics.
The increase in filter depth also compensates for the reduction in spatial resolution caused by pooling operations.

All convolutional layers use $3 \times 3$ kernels with \texttt{same} padding, preserving spatial dimensions and allowing the network to keep extracting features at the same resolution before downsampling is applied later via pooling.

Instead of using a flattening operation, we apply Global Average Pooling at the end of the convolutional stack. 
This reduces each feature map to a single value, drastically lowering the number of parameters.

The final classifier consists of two fully connected layers with dropout regularization, followed by a softmax output layer for multi-class prediction.

This architecture was achieved by trial and error until performance plateaued at a weighted F1-score of 0.58. 
Furthermore, we reached the conclusion that our model used mainly color as a discriminative feature (likely influenced by the limited number of training samples).
As a result, we decided to adopt transfer learning with pre-trained models in order to capture more complex features such as shapes, textures, and corners. 


\subsection{Transfer Learning with Pre-trained Models}
For all architectures described below, pretrained weights from ImageNet were used and the backbone networks were treated as fixed feature extractors, with their layers frozen during training. 
A custom classification head was added on top to adapt each model to the artist classification task.

\subsubsection{ResNet50}

ResNet50 is a deep convolutional neural network architecture composed of 50 layers, based on the concept of residual learning. 
Its key feature is the use of residual connections, which allow the network to learn residual mappings instead of direct transformations. 
These connections facilitate the learning of both low-level and high-level features, improving the model’s capacity to generalize across varying artistic styles and making ResNet50 a suitable choice for our problem.


\subsubsection{InceptionV3}

InceptionV3 is a convolutional neural network architecture designed to capture visual features at multiple scales. 
Its main idea is the use of Inception modules, where multiple operations (e.g., $1 \times 1$, $3 \times 3$, $5 \times 5$ convolutions and pooling) are applied in parallel to the same input and their outputs are concatenated. 
This allows the model to extract different types of features simultaneously.
This architecture is particularly useful when the size of relevant patterns is unknown, as it can capture both fine details and more global structures. 
For our task, this is beneficial since artistic styles can vary in both local features and overall composition.


\subsubsection{Vision Transformer (ViT)}

ViT is an architecture that applies transformer models to image classification. 
Instead of using convolutions, the image is divided into fixed-size patches, which are treated as a sequence and processed using self-attention mechanisms, allowing the model to capture global relationships between different parts of the image.
This is useful for artist classification, where overall composition and spatial relationships can be important.


\subsection{Fine-Tuning Strategy}
Focus on EfficientNet, frozen layers vs trainable layers, learning rates.

\section{Experimental Setup}

All models were trained using the Adam optimizer with a learning rate of $10^{-3}$ and the sparse categorical cross-entropy loss function, which is suitable for multi-class classification with integer-encoded labels.

Model performance was primarily evaluated using the F1-score, as it provides a better assessment under class imbalance, although accuracy was also monitored during training.

Training was performed for a maximum of 60 epochs. 
To prevent overfitting and reduce unnecessary computation, early stopping was applied based on the validation loss, with a patience of 10 epochs for CNN-based models and 5 epochs for the Vision Transformer. 
The best model weights were restored after training.

Additionally, a learning rate scheduler (ReduceLROnPlateau) was used to reduce the learning rate by a factor of 0.5 when validation loss plateaued, with a minimum learning rate of $10^{-6}$.
Model checkpoints were saved based on the best validation loss.

\section{Results and Evaluation}
\subsection{Quantitative Results}
F1-score, loss, comparison table.
\subsection{Model Comparison}
Discussion of performance differences, potential reasons.
\subsection{Error Analysis}
Common misclassifications, patterns.

\section{Discussion}
Interpretation of results, trade-offs between models, challenges (dataset size, overfitting, etc).

\section{Conclusion}
Summary of findings, possible improvements, future work.




\printbibliography

\end{document}

